---
title: 'Lab 2: Data wrangling'
output: html_document
---
In this first "real" lab we'll go over one of the most important skills in data science: Data Wrangling. It's the art of cleaning and manipulating our data to prepare it for our analyses and visualizations. A common estimate is that for most machine-learning projects, 80% of the time is spent on data wrangling, and the remaining 20% is spent on actually setting up and training the machine learning model.

We'll go over three main things today: basic data manipulation skills, web scraping, and loading and combining datasets.

We'll start with basic data manipulation skills. It is useful to FIRST try to visualize in our head what we want to do with the data, and THEN look for the code that can do that for us. Once you learn some useful functions, you will be able to combine them to manipulate the data to your will like a wizard.

You can find a list of our spells... ehhhm, functions [here](https://rstudio.com/wp-content/uploads/2015/02/data-wrangling-cheatsheet.pdf).


And a list of other useful spellbooks (cheat sheets) [here](https://rstudio.com/resources/cheatsheets/). 


### 1. Getting an overview of the data
First we'll explore different ways to get an overview of the dataset.
Let's use the gapminder dataset, available in the `dslabs` package. Because the data is available in the package, we can load it by simply using `data(gapminder)`. First we'll look at the dataset to see what it contains. We will use `summary()`, `head()` and `str()`.

##### 1) What type of information do each one of these functions provide about our dataset?

##### 2) How is the dataset ordered?

##### 3) What range of years does our dataset contain?

```{r, message=FALSE, first-plot}
# install.packages('dplyr')
# install.packages('dslabs')
# install.packages('tidyr')
library(tidyr)
library(dplyr)
library(dslabs)
library(rvest)
data(gapminder)

df <- gapminder

```


### 2. Ordering and subsetting our dataset

We can subset our dataset in different ways.

For example, the brute-force approach: we can select the cell in row 1, column 3 using `gapminder[1,3]`. We can also select a range of rows using `:`; i.e. we can select rows 2 to 5, column 1 using `gapminder[2:5,1]`.

We can use the `filter()` function to select rows according to values of a column, and we can use `select()` to select only specific columns according to their name.

We can also order a data frame using `arrange()`

##### 1) Create a data frame called `df2` that is ordered so that it shows the data for each country (all years), then the next country (all years), etc.

##### 2) Create a data frame called `algeriagdp` that includes only year and GDP data for Algeria.

##### 3) Think of 2 other ways we could have achieved the same result for #2. You can code them if you want, but the main task is to think of the process.


```{r}
### Selecting Columns
select(df, life_expectancy)
df$life_expectancy
df[,c('life_expectancy')]

### Filtering Rows
df[1:10, ] ## chooses rows 1-10
df[df$life_expectancy <= 70, ] ## filter based on condition

# df[some logical T/F, ]
# same as line 67 but split into 2 lines
condition_vector <- df$life_expectancy <= 70
df[condition_vector, ]

a <- filter(df, life_expectancy <= 70)
b <- df[df$life_expectancy <= 70, ]

### Multiple conditions
c <- filter(df, life_expectancy <= 70, year > 2000)
d <- df[df$life_expectancy <= 70 & df$year > 2000, ]


### Filter + Select Together
# "Crude Way"
df[df$life_expectancy <= 70 & df$year > 2000, c('population', 'country')]

# "Tidyverse Way" (More like English)
df1 <- filter(df, life_expectancy <= 70, year > 2000)
df2 <- select(df1, population, country)


### Creating new variables
## Task: Create rate of infant mortality / population
# Crude Way
### Neat shortcut for indent = CMD I or windows equivalent 
### (see Tools > Keyboard shortcut help)
df$infant_mortality_rate <-
  df$infant_mortality/df$population

df$infant_mortality_rate <-
  df$infant_mortality/df$population

# tidyverse way (mutate)
### same thing, second one more readable but both work
mutate(df, infant_mortality_rate = infant_mortality/population)

### Multiple columns at once
df3 <- mutate(df, 
              "infant_mortality_rate" = infant_mortality/population,
              "copy_of_country" = country)

### Sorts a data frame (arrange)
## by default sorts ascending
arrange(df, life_expectancy)

### make it descending
arrange(df, desc(life_expectancy))

### multiple columns
### First sort df by country (alphabetically)
### In the case of ties, sort by year
df_sorted <- arrange(df, country, year)
df_sorted2 <- arrange(df, country, desc(year))



```

```{r}
#1




#2


#3


```

### 2.1 (optional) The extremely useful `which()` function.

The `which()` function returns the index of rows in a data frame that meet a certain criteria. For example, if we want to know the index of the rows for Algeria, we can type the following:

```{r}
which(gapminder$country == "Algeria")
```

We can use this to select rows within our data frame by using brackets `[ ]` and placing our `which()` function where we would place the row number.
```{r}
head(gapminder[which(gapminder$country == "Algeria"), ])

# Note that R requires us to put a comma afterwards to indicate the column numbers. We can leave it blank to indicate we want all columns. We can also indicate the columns we want, either using the column number or by putting the column name in brackets, e.g. "infant_mortality"

head(gapminder[which(gapminder$country == "Algeria"), "infant_mortality"])
```

We can also use this to change the values of specific rows or cells. For example, Swaziland recently changed the country name to Eswatini. If we want to make this change, we can do the following:

```{r}
gapminder$country = as.character(gapminder$country)

gapminder$country[gapminder$country == "Swaziland"] = "Eswatini"

gapminder$country = as.factor(gapminder$country)
```



### 3. The pipe operator %>%

To generate the result above, you may have needed steps in between. We can use the pipe operator `%>%` to avoid using intermediate steps. This makes for much more intuitive and fun data wrangling.

For example, we can create the data frame mentioned above in an elegant, intuitive way:
```{r}
algeriagdp <- gapminder %>% 
  filter(country == 'Algeria') %>% 
  select (gdp,year)
```

##### 1) Using the pipe operator, create a dataset that includes country, year, gdp and population for countries in South-Eastern Asia between 1965 and 2010. Order the data frame by country, then year. Try to do it all in one command, starting from the `gapminder` dataset, using the pipe operator. Confirm that your code is doing what it's supposed to.

```{r}
#asiadf <- gapminder %>% #your code here

```

### 4. Creating variables

We can create variables using `mutate()`.

##### 1) Create a variable with the approximate number of women in the population. Assume that 52% of the population are women.

##### 2) Calculate an approximation of the number of children born each year based on the `fertility` column, which displays the expected number of children per woman over their reproductive life.
For this, assume an average reproductive life of 25 years.

##### 3) Look up and explore the function `lag()`. Use it to calculate the change in population from the year prior, and assign it to a column called `popchange`.

```{r}
#1

#2

#3


```


### 5. Summary

As we saw in class, the function `summarize()` provides a useful way to combine data from different rows.

##### 1) Create a dataset with the total world population by year.

##### 2) Create a dataset with the gdp per capita by world region for the year 2010. Be careful of the order when using the pipe operator `%>%`

##### 3) Calculate the increase in gdp per capita from the year prior by region and year. Explain any assumptions you had to make.

```{r}
#1

#2

#3

```


