## Data types

Variables in R can be of different types. For example, we need to distinguish numbers from character strings and tables from simple lists of numbers. The function `class` helps us determine what type of object we have:

```{r}
a <- 2
class(a)
```

To work efficiently in R it is important to learn the different types of variables and what we can do with these.

### Data Frames

Up to now, the variables we have defined are just one number. This is not very useful for storing data. The most common way of storing a dataset in R is in a _data frame_. Conceptually, we can think of a data frame as a table with rows representing observations and the different variables reported for each observatin defining the columns. Data frames are particularly useful for datasets because we can combine different data types into one object. 

We stored the data for our motivating example in a data frame. You can access this dataset by loading the `dslabs` library and loading the `murders` dataset using the `data` function:

```{r}
library(dslabs)
data(murders)
```

To see that this is in fact a data frame we type

```{r}
class(murders)
```

### Examining an object

The function `str` is useful to find out more about the structure of an object

```{r}
str(murders)
```

This tells us much more about the object. We see that the table has 51 rows (50 states plus DC) and five variables. We can show the first six lines using the function `head`:

```{r}
head(murders)
```

In this dataset each state is considered an observation and five variables are reported for each state.

Before we go any further in answering our original question about different states, let's get to know the components of this object better.

### The accessor

For our analysis we will need to access the different variables, represented by columns, included in this data frame. To access these variables we use the accessor operator `$` in the following way:

```{r}
murders$population
```

But how did we know to use `population`? Above, by applying the function `str` to the obejct `murders`, we revealed the names for each of the five variables stored in this table. We can quickly access the variables names using:

```{r}
names(murders)
```

It is important to know that the order of the entries in `murders$population` preserve the order of the rows in our data table. This will later permit us to manipulate one variable based on the results of another. For example, we will be able to order the state names by number of murders.

**Tip**: R comes with a very nice auto-complete functionality that saves us the trouble of typing out all the names. Try typing `murders$p` then hitting the _tab_ key on your keyboard. RStudio has many useful auto-complete feature options.

### Vectors: numerics, characters, and logical

Note that the object `murders$population` is not one number but several. We call these types of objects _vectors_. A single number is technically a vector but in general vectors refer to objects with several entries. The function `length` tells you how many entries are in the vector:

```{r}
pop <- murders$population
length(pop)
```

This particular vector is _numeric_ since population sizes are numbers:

```{r}
class(pop)
```

In a numeric vector, every entry must be a number. 

To store character strings, vectors can also be of class _character_. For example, the state names are characters:

```{r}
class(murders$state)
```

As with numeric vectors, all entries in a character vector need to be a character.

Another important type  are _logical vectors_. These must be either `TRUE` or `FALSE`. 

```{r}
z <- 3 == 2
z
class(z)
```

Here the `==` is a relational operator asking if 3 is equal to 2. Remember that in R, if you just use one `=` when you actually assign avalue. You can see the other _relational operators_ by typing

```{r, eval=FALSE}
?Comparison
```

In future sections you will see how useful relational operators can be.


### Factors

In the `murders` dataset we might expect the region to also be a character vector. However, it is not:

```{r}
class(murders$region)
```

it is a _factor_. Factors are useful for storing categorical data. Notice that there are only 4 regions:


```{r}
levels(murders$region)
```

So, in the background, R stores these _levels_ as integers and keeps a map to keep track of the labels. This is more memory efficient than storing all the characters. However, factors are also a source of confusion as they can easily be confused with characters but behave differently in different contexts. We will see more of this later. 

In general, we recommend avoiding factors as much as possible although they are sometimes necessary to fit models containing categorical data.
 
### Lists

Data frames are a special case of _lists_. We will covers lists in more detail later but know that they are useful because you can store any combination of other types. Here is an example of a list we created for you:


```{r}
record <- list(name = "John Doe",
               student_id = 1234,
               grades = c(95, 82, 91, 97, 93),
               final_grade = "A")
```

```{r}
record
class(record)
```

We won't be using lists until later but you might encounter one in your own exploration of R. Note that, as with data frames, you can extract the components with the accessor `$`. In fact, data frames are a type of list.

```{r}
record$student_id
```

We can also use double brackets like this:

```{r}
record[["student_id"]]
```

You should get used to the fact that in R there are several ways to do the same thing, in particular accessing entries.

---

## Vectors

The most basic unit available in R to store data are _vectors_. Complex datasets can usually be broken down into components that are vectors. For example, in a data frame, each column is a vector. Here we learn more about this important class.

### Creating vectors

We can create vectors using the function `c`, which stands for concatenate. We use `c` to _concatenate_ entires in the following way:

```{r}
codes <- c(380, 124, 818)
codes
```

We can also create character vectors. We use the quotes to denote that the entries are characters rather than variables names.

```{r}
country <- c("italy", "canada", "egypt")
```

Note that if you type

```{r, eval=FALSE}
country <- c(italy, canada, egypt)
```

you recieve an error becuase the variables `italy`, `canada` and `egypt` are not defined: R looks for variables with those names and returns an error.

### Names

Sometimes it is useful to name the entries of a vector. For example, when defining a vector of country codes we can use the names to connect the two:

```{r}
codes <- c(italy = 380, canada = 124, egypt = 818)
codes
```

The object `codes` continues to be a numeric vector:
```{r}
class(codes)
```

but with names
```{r}
names(codes)
```

If the use of strings without quotes looks confusing, know that you can use the quotes as well

```{r}
codes <- c("italy" = 380, "canada" = 124, "egypt" = 818)
codes
```

There is no difference between this call and the previous one: one of the many ways R is quirky compared to other languages.

We can also assign names using the `names` function:

```{r}
codes <- c(380, 124, 818)
country <- c("italy","canada","egypt")
names(codes) <- country
codes
```

### Sequences

Another useful function for creating vectors generates sequences

```{r}
seq(1, 10)
```

The first argument defines the start, and the second the end. The default is to go up in increments of 1, but a third argument let's us tell it how much to jump by:

```{r}
seq(1, 10, 2)
```

If we want consecutive integers we can use the following shorthand

```{r}
1:10
```

Note that when we use this function, R produces integers, not numerics, because they are typically used to index something:

```{r}
class(1:10)
```

However, note that as soon as we create something that's not an integer the class changes:

```{r}
class(seq(1, 10))
class(seq(1, 10, 0.5))
```

### Subsetting

We use square brackets to access specific elements of a list. For the vector `codes` we defined above, we can access the second element using
```{r}
codes[2]
```

You can get more than one entry by using a multi-entry vector as an index:
```{r}
codes[c(1,3)]
```

The sequences defined above are particularly useful if we want to access, say, the first two elements

```{r}
codes[1:2]
```

If the elements have names, we can also access the entries using these names. Here are two examples.

```{r}
codes["canada"]
codes[c("egypt","italy")]
```

### Coercion

In general, _coercion_ is an attempt by R to be flexible with data types. When an entry does not match the expected, R tries to guess what we meant before throwing an error. This can also lead to confusion. Failing to understand _coercion_ can drive a programmer crazy when attempting to code in R since it behaves quite diffently from most other languages in this regard. Let's learn about it with some examples.

We said that elements of a vector must be all of the same type. So if we try to combine, say, numbers and characters you might expect an error

```{r}
x <- c(1, "canada", 3)
```

But we don't get one, not even a warning! What happened? Look at `x` and its class:

```{r}
x
class(x)
```

R _coerced_ the data into characters. It guessed that because you put a character string in the vector you meant the 1 and 3 to actually be character strings `"1"` and `"3"`. The fact that not even a warning is issued is an example of how coercion can cause many unnoticed errors in R. 

R also offers functions to force a specific coercion. For example you can turn numbers into characters with 

```{r}
x <- 1:5
y <- as.character(x)
y
```

And you can turn it back with `as.numeric`. 

```{r}
as.numeric(y)
```
 
This function is actually quite useful as datasets that include numbers as character strings are common. 

### Not Availables (NA)

When these coercion functions encounter an impossible case it gives us a warning and turns the entry into a special value called an `NA` for "not available".  For example:

```{r}
x <- c("1", "b", "3")
as.numeric(x)
```

R does not have any guesses for what number you want when you type `b` so it does not try.

Note that as a data scientist you will encounter `NA`s often as they are used for missing data, a common problem in real-life datasets.




