## Assessments


1. What will this conditional expression return?

    ```{r}
    x <- c(1,2,-3,4)
    
    if(all(x>0)){
      print("All positives")
    } else{
      print("Not all positives")
    }
    ```

"Not all positives", because there is at least one element of `x` that is not positive, namely the 3rd one.

2. Which of the following expressions is always `FALSE` when at least one entry of a logical vector `x` is TRUE?

    A. `all(x)`
    B. `any(x)`
    C. `any(!x)`
    D. `all(!x)`

D. If at least one entry of `x` is `TRUE`, then at least one entry of `!x` is `FALSE`. Therefore, `all(!x)` is `FALSE` since there is at least one entry of `!x` that is `FALSE`.

Here are counterexample for the other three choices:

```{r}
# counter-example for A that returns TRUE
x <- c(TRUE, TRUE)
all(x)
    
# counter-example for B and C that return TRUE
x <- c(TRUE, FALSE)
any(x)
any(!x)
```

3. The function `nchar` tells you how many characters long a character vector is. For example:

    
```{r}
library(dslabs)
data(murders)
char_len <- nchar(murders$state)
char_len[1:5]
```


Write a line of code that assigns to the object `new_names` the state abbreviation when the state name is longer than 8 characters.

```{r}
new_names <- ifelse(nchar(murders$state) > 8,
                    murders$abb,
                    murders$state)
new_names
```   

